---
title: "Halloween Mini-Project"
author: "Mariam Benny (A17103451)"
format: pdf
---

```{r}
candy <- read.csv("candy-data.csv", row.names=1)
head(candy)
```


>Q1. How many different candy types are in this dataset?

There's 85 candy types.
```{r}
nrow(candy)
```

>Q2. How many fruity candy types are in the dataset?

There's 38 fruity candy types.

```{r}
sum(candy$fruity)
```

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy is Rolo, and its winpercent is 65.7%.

```{r}
candy["Rolo","winpercent"]
```

>Q4. What is the winpercent value for “Kit Kat”?

Kit Kat has a 76.8% winpercent.

```{r}
candy["Kit Kat",]$winpercent
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

Tootsie Roll Snack Bars has a 49.7% winpercent.

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
library(dplyr)

candy |>
  filter(rownames(candy)=="Haribo Happy Cola") |>
  select(winpercent)
```

>Q. Find a fruit candy with a winpercent above 50%.

```{r}
candy |>
  filter(winpercent > 50) |>
  filter(fruity==1)
```


```{r}
top.candy <- candy[candy$winpercent > 50,]
top.candy[top.candy$fruity ==1,]
```


To get a quick insight into a new data set
```{r}
library("skimr")
skim(candy)
```

```{r}
skimr::skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

`winpercent` variable looks different, its range is much larger than just 0 to 1.

>Q7. What do you think a zero and one represent for the candy$chocolate column?

It likely represents whether chocolate is present in that candy or not, 0 meaning not present and 1 meaning it is.
```{r}
candy$chocolate
```

>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent, breaks=10)
```

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(binwidth=8) +
  theme_classic()
```


>Q9. Is the distribution of winpercent values symmetrical?

No it is not symmetrical, it skews to one side.

>Q10. Is the center of the distribution above or below 50%?

Below 50%, the center median is 47.83%.

```{r}
summary(candy$winpercent)
```

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Chocolate candy is higher ranked (60.8 median) versus fruit candy (43.0 median).

```{r}
fruit.candy <- candy |>
  filter(fruity==1)

summary(fruit.candy$winpercent)
```
```{r}
#summary(candy[as.logical(candy$chocolate),]$winpercent)
chocolate.candy <- candy |>
  filter(chocolate==1)

summary(chocolate.candy$winpercent)
```

>Q12. Is this difference statistically significant?

Yes, p-value = 2.871e-08, it's less than .05 so that makes it significant.
```{r}
t.test(chocolate.candy$winpercent, fruit.candy$winpercent)
```


>Q13. What are the five least liked candy types in this set?

Bottom 5:

* Nik L Nip
* Boston Baked Beans
* Chiclets
* Super Bubble
* Jawbusters

```{r}
head(candy[order(candy$winpercent),],5)
```

>Q14. What are the top 5 all time favorite candy types out of this set?

Top 5:

* Snickers
* Kit Kat
* Twix
* Reese's Miniatures
* Reese's Peanut Butter cup

```{r}
tail(candy[order(candy$winpercent),],5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(x=winpercent, 
      y=reorder(rownames(candy), winpercent),
      fill=chocolate) +
  geom_col()
```

I want a more custom color scheme where I can see both chocolate and bar and fruity etc. all from one plot. To do this we can roll our own  color vector...

```{r}
mycols <- rep("black",nrow(candy))
mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols[as.logical(candy$bar)] <- "brown"
mycols[as.logical(candy$fruity)] = "pink"
```

```{r}
mycols[rownames(candy)=="Rolo"] <- "blue"
```


```{r}
ggplot(candy) +
  aes(x=winpercent, 
      y=reorder(rownames(candy), winpercent)) +
  geom_col(fill=mycols)
```


>Q17. What is the worst ranked chocolate candy?

Worst ranked chocolate candy is Sixlets.

>Q18. What is the best ranked fruity candy?

Best ranked fruity is Starbursts.

Plot of winpercent vs pricepercent to see what would be the best candy to buy

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent) +
  geom_point(col=mycols) 
```

```{r}
library(ggrepel)

mycols[as.logical(candy$fruity)] = "purple"

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, size=3.3, max.overlaps = 5)
```

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) + 
  geom_point(col=mycols) +
  geom_text_repel(col=mycols, max.overlaps = 8)
```


>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's minuatures


>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The top 5 most expensive and least popular are:

* Nik L Nip
* Nestle Smarties
* Ring pop
* Hershey's Krackel
* Hershey's Milk Chocolate

The least popular is Nik L Nip.

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

```{r}
library(corrplot)

```

```{r}
cij <- cor(candy)
corrplot(cij)
```


>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruit and chocolate are anti-correlated.


>Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent are most positively correlated.


# Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2], col=mycols, pch=16)
```

How do the original variables contribute to the new PCs. I will look at PC1 here.

```{r}
pca$rotation
```

```{r}
loadings <- as.data.frame(pca$rotation)

ggplot(loadings) + 
  aes(PC1, reorder(rownames(loadings), PC1), fill=PC1) +
  geom_col()

```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, pluribus, and hard are variables that are picked up strongly by PC1 in the positive direction. Yes, this makes sense to me because the variables are correlated to each other among the most popular candies.

```{r}
my_data <- cbind(candy, pca$x[,1:3])

```


```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=mycols)

p
```



```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```





